Do larger watersheds respond different to forest cover change? Re-analysing a global data set.

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.tex
updates on paper 12012022

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

  \title{Do larger watersheds respond different to forest cover change? Re-analysing a global data set.}
    \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi\corref{2}}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso\corref{2}}
   \ead{jalonso@fing.edu.uy} 
      \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
    \address[INIA]{Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  This is the abstract.
  It consists of two paragraphs.
  \end{abstract}

@@

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017). The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in Zhang et al. (2011) indicating (for Australian watersheds) a 100\% decrease in streamflow for watersheds with 100\% forest cover. However, on the other end of the spectrum, in a series of French watersheds (Cosandey et al., 2005), there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.

Several review papers have summarized different studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017). These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover. The Zhang et al. (2017) dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by Filoso et al. (2017) focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.

@@

The second study (Filoso et al., 2017) was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a relationship between watershed size and decline in streamflow.

@@

Encouraged by the work presented by Zhang et al. (2017) and Filoso et al. (2017) and the fantastic database of studies presented by these authors, we believe we can add to the discussion. In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.

In particular, the main method in the work by Zhang et al. (2017) is using simple linear regression, and in Filoso et al. (2017) the focus is mainly on classification. As Zhang et al. (2017) points out, the main assumption in their work is that the threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' watersheds, but the subset of data in Filoso et al. (2017) does not appear to support this. And while te work Filoso et al. (2017) provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact. This is because the work had a strict criterion to select quantitative studies. However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km\textsuperscript{2}.

As a result the objective of this paper is to 1) enhance the data set from Zhang et al. (2017) with further watersheds (such as from Filoso et al. (2017)) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).

Building on the analyses by Zhang et al. (2017) and Filoso et al. (2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths Zhou et al. (2015)), the length of the study (c.f. Jackson et al. (2005)) and possibly the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al. (2017); Zhou et al. (2015)).

@@

While Zhang et al. (2017) use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) watershed data sets in our analysis.

@@

Several additional data points from watershed studies were extracted from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988), Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010), Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al. (2011)).
We also removed one data point from the analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al. (2017). This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.

@@

Removing possible duplicates removes 29 watersheds from the overall data set

@@

\[\tag{2}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}\]

@@

The overall dataset contains 309 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from Zhang et al. (2017) contained 312 observations. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42\% of the data from watersheds \textless{} 1 km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10 km\textsuperscript{2}.

\begin{figure}
\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{foo}\label{fig:datagraphs}
\end{figure}

@@

\[\tag{3}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}\]

@@

\[\tag{4}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
\end{aligned}\]

@@

\[\tag{5}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
\end{aligned}\]

@@

The work by Filoso et al. (2017) and earlier by Jackson et al. (2005) has indicated that the length of the study might influence the response. This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al. (2005)). In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 17.09 & 6.08 & 2.81 & 0.01 \\
\textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
\textbf{Forest\_Signincrease} & -34.6 & 6.11 & -5.67 & 0 \\
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
\caption{Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule
\endhead
\textbf{s(log10(Area\_km2))} & 1.74 & 9 & 0.78 & 0.01 \\
\textbf{s(Dryness)} & 0.87 & 9 & 0.57 & 0.02 \\
\textbf{s(length)} & 8.83 & 9 & 3.1 & 0 \\
\bottomrule
\end{longtable}

Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.29 and deviance explained of 0.32, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}). However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.

A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.

>>>>

@@

  \title{Do larger watersheds respond different to forest cover change? Re-analysing a global data set.}
    \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso}
   \ead{jalonso@fing.edu.uy} 
      \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables.
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data \textless{} 1 km2 and 65\% of the data \textless{} 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38/\%) is low due the quality of the inputs and additional unknown confounding factors.
  \end{abstract}

@@

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017). The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in Zhang et al. (2011) indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover. However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchmensts after deforestation.

Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017). These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover. The Zhang et al. (2017) dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by Filoso et al. (2017) focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets.

@@

The second study (Filoso et al., 2017) was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a log-linear relationship between decreasing watershed size and an increasing decline in streamflow. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

@@

Encouraged by the work presented by Zhang et al. (2017), Filoso et al. (2017) and Zhou et al. (2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion. In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.

In particular, the main method in the work by Zhang et al. (2017) is a single covariate linear regression, and in Filoso et al. (2017) the focus is mainly on classification and there is again some single covariate linear regression. As Zhang et al. (2017) points out, a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments. However, the subset of 37 data points in Filoso et al. (2017) (their Figure 9) does not appear to support this, suggesting a continuum. And while the work Filoso et al. (2017) provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.

As a result the objective of this paper is to 1) enhance the data set from Zhang et al. (2017) with further catchments (such as from Filoso et al. (2017)) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).

Building on the analyses by Zhang et al. (2017) and Filoso et al. (2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths Zhou et al. (2015)), the length of the study (c.f. Jackson et al. (2005); Filoso et al. (2017)) and the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al. (2017); Zhou et al. (2015)).

@@

While Zhang et al. (2017) use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) catchment data sets in our analysis.

@@

Several additional data points from watershed studies were extracted from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988), Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010), Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al. (2011)).
We also removed one data point from the analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al. (2017). This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for watershed \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984). Finally, on review of all the data in Zhang et al. (2017) and Filoso et al. (2017), 29 potential duplicates were identified and flagged in the data.

@@

@@

\[\tag{2}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}\]

@@

The overall dataset contains 309 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from Zhang et al. (2017) contained 312 watersheds and the Filoso et al. (2017) study used 37 watersheds (Table S2 in Filoso et al. (2017)). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42\% of the data from watersheds \textless{} 1 km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10 km\textsuperscript{2}.

\begin{figure}
\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables.}\label{fig:datagraphs}
\end{figure}

@@

\[\tag{3}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}\]

@@

\[\tag{4}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
\end{aligned}\]

@@

\[\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
\end{aligned}\]

@@

The work by Filoso et al. (2017) and earlier by Jackson et al. (2005) has indicated that the length of the study might influence the response. This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al. (2005)). In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (\emph{Dryness, Area, From}) were considered non-linear in this model. This is also indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables. \emph{From} represents the starting date of the data collection.

\[\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
\end{aligned}\]

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 17.37 & 5.91 & 2.94 & 0 \\
\textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
\textbf{Forest\_Signincrease} & -34.93 & 5.84 & -5.98 & 0 \\
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
\caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule
\endhead
\textbf{s(log10(Area\_km2))} & 2.48 & 9 & 0.55 & 0.1 \\
\textbf{s(Dryness)} & 0.89 & 9 & 0.62 & 0.01 \\
\textbf{s(length)} & 0 & 9 & 0 & 0.66 \\
\textbf{s(From)} & 8.1 & 9 & 3.97 & 0 \\
\bottomrule
\end{longtable}

Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.31 and deviance explained of 0.34, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}). However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.

A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.42}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
\caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & -21.06 & 19.16 & -1.1 & 0.27 \\
\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 2.99 & 0 \\
\textbf{Forest\_Signincrease} & -23.11 & 7.13 & -3.24 & 0 \\
\textbf{Precip\_data\_typeOB} & -7.65 & 14.51 & -0.53 & 0.6 \\
\textbf{Precip\_data\_typeSG} & 18.92 & 16.17 & 1.17 & 0.24 \\
\textbf{Assessment\_techniqueEA, HM} & 17.73 & 45.71 & 0.39 & 0.7 \\
\textbf{Assessment\_techniqueHM} & 29.82 & 12.66 & 2.36 & 0.02 \\
\textbf{Assessment\_techniquePWE} & 44.69 & 12.68 & 3.53 & 0 \\
\textbf{Assessment\_techniquePWE,
HM} & 39.72 & 46.79 & 0.85 & 0.4 \\
\textbf{Assessment\_techniqueQPW} & 37.54 & 21.54 & 1.74 & 0.08 \\
\textbf{Assessment\_techniqueQPW,
EA} & 24.9 & 27.45 & 0.91 & 0.37 \\
\textbf{Assessment\_techniqueSH} & 43.68 & 13.27 & 3.29 & 0 \\
\textbf{Forest\_typeCF} & -4.06 & 8.71 & -0.47 & 0.64 \\
\textbf{Forest\_typeMF} & -8.33 & 8.7 & -0.96 & 0.34 \\
\textbf{Hydrological\_regimeSD} & 9.41 & 10.57 & 0.89 & 0.37 \\
\bottomrule
\end{longtable}
